---
title: "Assignment 2- OX"
author: "Owen Xu"
date: "04/09/2019"
output:
  html_document: default
  pdf_document: default
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


###Question 1
This question works with a data set on sodium intake.

```{r }
##Read data
sodium <- read.table("sodium.txt", header=TRUE)

##Check the first few rows of data          
head(sodium)

##get unique combination of 1st and 2nd columns
unique(sodium[,c("Instructor","Supplement")])

##get the index as numeric values 
ind = as.numeric(rownames(unique(sodium[,c("Instructor","Supplement")])))
                 
##find the Sodium values
data = sodium[ind,"Sodium"]

##create the matrix by row
sodiumMat = matrix(data,ncol = 4,byrow = TRUE)

##name the matrix
colnames(sodiumMat) = c("A","B","C","D")
rownames(sodiumMat) = unique(sodium[,"Instructor"])
names(dimnames(sodiumMat)) <- c("Instructor","Supplement")

##show the matrix
sodiumMat

```

fit the model:


```{r question 1}
##get total mean miu and sweep 
miu = mean(sodiumMat)
sodiumMat = sweep(sodiumMat,2, miu,FUN = "-")

##get the row mean alpha
alpha = apply(sodiumMat,1,mean)

##get the column mean beta
beta = apply(sodiumMat,2,mean)

##sweep both values
sodiumMat = sweep(sodiumMat,1,alpha,FUN = "-")
sodiumMat = sweep(sodiumMat,2,beta,FUN = "-")

##show final fitted matrix
sodiumMat

```
This does not look like the model that works because the residuals can be quite large for each instructor.


###Question 2
This question works with a set of plant weights, measured under two experimental conditions.

```{r}
## Annette Dobson (1990) "An Introduction to Generalized Linear Models".
## Page 9: Plant Weight Data.
## Control = standard conditions
## Treatment = nutrient rich
ctl <- c(4.17,5.58,5.18,6.11,4.50,4.61,5.17,4.53,5.33,5.14)
trt <- c(4.81,4.17,4.41,3.59,5.87,3.83,6.03,4.89,4.32,4.69)
group <- gl(2, 10, 20, labels = c("Ctl","Trt"))
weight <- c(ctl, trt)
```

We will first assume that all weights are i.i.d. Normal(μ, σ). We will further assume that σ is the sample standard deviation.
```{r}
sigma <- sd(weight)

```
##like function that calculates the likelihood
```{r}
like <- function(x,mean){
  prod(dnorm(x,mean,sigma))
}

##test
like(weight,0)
like(weight,4)
```


```{r}
##plot mu = 0

par(mfrow = c(1,2))
plot(weight,dnorm(weight,0,sigma),ylab = "density", type="h",main = expression(mu==0))

##plot mu = 4
plot(weight,dnorm(weight,4,sigma),ylab = "density", type="h",main = expression(mu==4))

##plot likelihood for mu = 0 - 6
##assign 51 data points between 0 and 6
x = seq(0,6,length = 51)
len = length(x)
likevec = numeric(len)

##get likelihood values into a vector for plot
for( i in 1:len)
  likevec[i] = like(weight,x[i])

##plot
plot(seq(0,6,length = 51),likevec, type="l",xlab = expression(mu),ylab = "Likelihood")

```
##log likelihood function
```{r}
loglike <- function(x,mean){
  sum(dnorm(x,mean,sigma,log= TRUE))
}

#tests
loglike(weight,0)
loglike(weight,4)
```
##plot

```{r}

par(mfrow = c(1,2))
##plot mu = 0
plot(weight,dnorm(weight,0,sigma, log = TRUE),ylab = "log density", type="h",main = expression(mu==0))

##plot mu = 4
plot(weight,dnorm(weight,4,sigma, log = TRUE),ylab = "log density", type="h",main = expression(mu==4))

##plot log likelihood for mu = 0 - 6
##assign 51 data points between 0 and 6
x = seq(0,6,length = 51)
len = length(x)
loglikevec = numeric(len)

##get likelihood values into a vector for plot
for( i in 1:len)
  loglikevec[i] = loglike(weight,x[i])

##plot
plot(seq(0,6,length = 51),loglikevec, type="l",xlab = expression(mu),ylab = "Log-Likelihood")

```
Use the optimise function to find the maximum likelihood estimate of μ (find the value
of μ that maximises the log-likelihood function).
```{r}
muMLE = optimise(loglike, c(0,6),x = weight, maximum = TRUE)

##muMLE equals mean of weight
muMLE == mean(weight)

```
Write a function dllike that calculates this 1st derivative and use uniroot to find where this function is zero.

```{r}
##dllike function
##dllike <- function (m) deriv(loglike(weight,m),"m")
dllike <- function(mean) sum(weight) - length(weight)*mean
uniroot(dllike,lower = -10,upper = 10,tol = 1e-20)

```
###Question 3
This question also works with the set of plant weights and we will still assume that all
weights are i.i.d. Normal(μ, σ).
However, we will now estimate both μ and σ using maximum likelihood.
```{r loglike2}
par(mfrow = c(2,2))

##define function loglike2
loglike2 <- function(x,mean,sigma){
  
  ##Recycle
  n = length(x)

  maxlen = length(mean)
  
  sigma = rep(sigma,maxlen)
  
  ##Vector calcs
  ans = dnorm(rep(x,maxlen),rep(mean,each = n),rep(sigma, each = n),log= TRUE)

  dim(ans) = c(n,maxlen)
  
  colSums(ans)
  
}

#evaluate (0,1) and (4,1)
loglike2(weight,0,1)
loglike2(weight,4,1)

#plot
plot(weight,dnorm(weight,0,1, log = TRUE),ylab = "log density", type="h",main = expression(paste(mu == 0,";",sigma == 1)))
plot(weight,dnorm(weight,4,1, log = TRUE),ylab = "log density", type="h",main = expression(paste(mu == 4,";",sigma == 1)))

#plot mu ~ (0 - 6) for sigma = 1 and 0.5
plot(seq(0,6,length = 51),loglike2(weight,mean = seq(0,6,length=51),sigma = 1), type="l",xlab = expression(mu),ylab = "Log-Likelihood", main = expression(sigma == 1))
plot(seq(0,6,length = 51),loglike2(weight,mean = seq(0,6,length=51),sigma = 0.5), type="l",xlab = expression(mu),ylab = "Log-Likelihood", main = expression(sigma == 0.5))



```


Use the optim function to find the maximum likelihood estimates for μ and σ.

```{r}
##rewrite loglike2 function
loglike2 <- function(x,theta){
  
sum(dnorm(x, theta[1], theta[2], log=TRUE))
  
}

##use optim;by observation from the above figure (5,1) is  a good starting points.
muSigmaMLE = optim(c(5,1),loglike2,x = weight, control=list(fnscale = -1))

muSigmaMLE$par

##compare
mean(weight)
sd(weight)
sd(weight)*sqrt((length(weight) - 1)/length(weight))

```
###Question 4
This question also works with the set of plant weights, but now we will allow there to be
a separate mean for the treatment and control groups.

```{r}

par(mfrow = c(2,2))

loglike3 <- function(x,g,b0,b1,sigma) {
  
  
  ##Recycle
  n = length(x)

  maxlen = length(b0)
  
  mean = rep(b0,each = n) + rep(g,maxlen) * b1 

  print(maxlen)
  
  sigma = rep(sigma,maxlen)
  
  ##Vector calcs
  ans = dnorm(rep(x,maxlen),mean,rep(sigma, each = n),log = TRUE)

  dim(ans) = c(n,maxlen)
  
  print(colSums(ans))
  
}

##calculate
 gp <- as.numeric(group) - 1
 
 loglike3(weight, gp, 0, 1, 1)

 loglike3(weight, gp, 4, 1, 1)

##plot
 
 mean1 = 0 + gp * 1
 plot(weight,dnorm(weight,mean1,1,log = TRUE),ylab = "log density", type="h",main = expression(paste("b0 = 0",";","b1 = 1",";",sigma == 1)))
 
 mean2 = 4 + gp * 1
 
 plot(weight,dnorm(weight,mean2,1,log = TRUE),ylab = "log density", type="h",main = expression(paste("b0 = 4",";","b1 = 1",";",sigma==1)))

#plot b0 ~ (0 - 6) for b1=1,sigma=1 and b1=0,sigma=1

 plot(seq(0,6,length = 51),loglike3(weight,b0 = seq(0,6,length = 51),g = gp,b1 = 0,sigma = 1), type="l",xlab = "b0",ylab =   
 "Log-Likelihood", main = expression(paste("b1 = 1",";",sigma == 1)))
 
 plot(seq(0,6,length = 51),loglike3(weight,b0 = seq(0,6,length = 51),g = gp,b1 = 1,sigma = 1), type="l",xlab = "b0",ylab = 
 "Log-Likelihood", main = expression(paste("b1 = 0",";",sigma == 1)))
 
 
```

```{r}
##rewrite loglike3 function
loglike3 <- function(dat,g,theta){

#gp <- as.numeric(group) - 1

mean = theta[1] + g * theta[2]
  
sum(dnorm(dat, mean, theta[3], log = TRUE))
  
}

##optimize;by observation from the above figure (4.5,0,1) is a good starting points.

params = optim(par = c(4.5,0,1), function(x) loglike3(weight, gp, x) ,control = list(fnscale = -1))
params$par

##compare
lm.D9 <- lm(weight ~ group)
coef(lm.D9)

```



